---
title: "Foundations of Probability and Statistics, Project"
author: "Emiliano Capasso, Antonello Scarcella, Simone Bellavia"
date: '2022-12-05'
geometry: "left=2cm,right=2cm,top=0.5cm,bottom=1.5cm"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      tidy.opts=list(width.cutoff = 80),
                      tidy = TRUE)
library(tidyverse) # for tidyverse
library(caret) # for createDataPartition
library(performance) # for compare_performance
library(ggplot2) # for plots
library(reshape2) # for melt function
library(gridExtra) # to arrange plots in a grid 
library(factoextra) # for PCA
library(corrplot) # for correlation between variables and PCA
```

# Introduction to Analysis

TBD

# Descriptive Analysis

TBD

## Importing data

```{r}
# import data
data <- read.csv("data.csv",
                    header = TRUE)
```

## First exploration

```{r}
# get structure of dataset
str(data)
```

- id: id number
- diagnosis (response): the diagnosis of breast tissues (M = malignant, B = benign)
- radius_mean: mean of distances from center to points on the perimeter
- texture_mean: standard deviation of gray-scale values
- perimeter_mean: mean size of the core tumor
- area_mean: //
- smoothness_mean: mean of local variation in radius lengths
- compactness_mean: mean of perimeter^2 / area - 1.0
- concavity_mean: mean of severity of concave portions of the contour
- concave points_mean: mean for number of concave portions of the contour
- ecc.

```{r}
# get summary of variables
summary(data)
```

```{r}
# getting the head of dataset
head(data)
```

## Missing Values

Checking if there are missing values:

```{r}
# check for missing values
colSums(is.na(data))
```

There aren't missing values in the considered dataset, except for 32th feature 'X' that is full of NA, we drop it.

```{r}
data <- data %>% select(-X)
```

We also drop id because it is not needed

```{r}
data <- data %>% select(-id)
```

Checking if the column has been dropped.

```{r}
colnames(data)
```

## Second exploration

```{r}
#describe(data)
```

Let's see the frequency of cancer diagnosis (if benignant or malignant)

```{r}
ggplot(data, aes(x = diagnosis)) + 
  geom_bar(aes(y = (after_stat(count))/sum(after_stat(count)))) + 
  scale_fill_manual(values = c("#0468BF","#D9A23D")) + 
theme_bw() +
  labs(x = "Diagnosis", y = "Frequency", title = "Frequency of Cancer Stages")
```

Benignant cancers are way more than malignant.

Converting the diagnosis values in numeric. Malignant == 1, Benign == 0.

```{r}
data$diagnosis <- as.numeric(data$diagnosis == "M")

str(data)
```

Let's see the correlation map with a heatmap.

```{r}
# Create correlation matrix
corr_matrix <- cor(data)

# Create heatmap
melted_matrix <- melt(corr_matrix)
ggplot(data = melted_matrix, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1,1), name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Correlation Map")
```

Features vs Diagnosis

```{r}
features_mean <- names(data)[2:12]

plots <- lapply(1:length(features_mean), function(x) {
  g <- ggplot(data, aes_string(x = features_mean[x], 
                              fill = as.factor(data$diagnosis))) +
    geom_histogram(binwidth = (max(data[,features_mean[x]]) - min(data[,features_mean[x]]))/50, 
                  alpha = 0.5, aes(color = as.factor(data$diagnosis))) +
    scale_fill_manual(values = c("green", "red")) +
    scale_color_manual(values = c("green", "red")) +
    ggtitle(features_mean[x]) +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(fill = "Diagnosis", color = "Diagnosis")
  return(g)
})

grid.arrange(grobs = plots, ncol = 2)
```

This code plots histograms of 10 features of a dataset with a binary diagnosis outcome. 
The histograms are stacked and show the distribution of each feature for each diagnosis type (red for "M" and green for "B"). 
The histograms are plotted in a 5x2 grid, with each feature in a separate plot and labeled with the feature name. 
The code uses ggplot2 library in R for plotting and gridExtra library for arranging the plots in a grid.
The resulting plot is a comparison of the distributions of each feature between the two diagnosis groups.

Mean values of texture, smoothness, symmetry or fractual dimension does not show a particular preference of one diagnosis over the other. 

## Reducing dimensionality

```{r}
# Exclude the response variable 'Diagnosis'
data_cont <- data[, which(colnames(data) != "diagnosis")]

# Scale th°e continuous features (if necessary)
data_cont_scaled <- scale(data_cont)

# Perform PCA on the scaled data
pca <- prcomp(data_cont_scaled)

# Summarize the results
summary(pca)

# Plot the first two principal components
plot(pca$x[,1:2], col=ifelse(data$Diagnosis==0, "blue", "red"),
     main="PCA of Continuous Features", xlab="PC1", ylab="PC2")
legend("topright", legend=c("B", "M"), col=c("blue", "red"), pch=1)
```

getting variables
```{r}

all_var <- get_pca_var(pca)
all_var

```

plotting correlations
```{r}

corrplot(all_var$cos2, is.corr = FALSE)

```

##

# Tests

we want to estimate the difference between the mean of radius size of all benign and all malignant
tumor to understand if the difference is real of given by sampling error. To do so, we draw two samples, one from the population of malignant tumor and another
from the population of benign tumor. These two samples are independent because they
are drawn from two different populations, and the samples have no effect on each other.

```{r}
mdf <- data[data$diagnosis == "B", ] # group of Malignant tumor
bdf <- data[data$diagnosis == "M", ] # group of Benign tumor
# Density Plot
dp <- ggplot(df, aes(x=radius_mean, group=diagnosis,fill=diagnosis)) +
geom_density(alpha=0.5) +
scale_fill_manual(values = c("#0468BF","#D9A23D")) + 
theme_bw()

dp
```

prior to run test on difference on mean, let's test
the normality first plotting the graph
```{r}
par(mfrow=c(1,2))
qqnorm(mdf$radius_mean, main = "Radius Mean distribution")
qqline(mdf$radius_mean)
qqnorm(bdf$radius_mean, main = "Radius Mean distribution")
qqline(bdf$radius_mean)
```

then using the Shapiro-Wilk normality test

```{r}
shapiro.test(mdf$radius_mean)
shapiro.test(bdf$radius_mean)
```

the p-value for the Benign group is > 0.05 so we accept the hypotesis
of normality for the group, whilst the malignant does not follow normality

we can try to remove the outliars and test again for normality

```{r}
# Calculate the IQR
IQR <- IQR(mdf$radius_mean)

# Calculate the lower and upper bounds
lower_bound <- quantile(mdf$radius_mean, 0.25) - 1.5 * IQR
upper_bound <- quantile(mdf$radius_mean, 0.75) + 1.5 * IQR

# Remove outliers
data_df_without_outliers <- mdf %>%
  filter(mdf$radius_mean >= lower_bound & mdf$radius_mean <= upper_bound)
```

we want to evaluate if the new sample is still acceptable
```{r}
boxplot(mdf$radius_mean,data_df_without_outliers$radius_mean,bdf$radius_mean, main="M/MTumor", ylab = "Tumor mean Radius (mm)")
```

then we test again for normality
```{r}
shapiro.test(data_df_without_outliers$radius_mean)
```

now we can make a test to evaluate the difference of the mean
```{r}
t.test(bdf$radius_mean,data_df_without_outliers$radius_mean, alternative="two.sided", var.equal=FALSE,conf.level=0.95)
```

p value is smaller than 0.05 so we can reject the null hypotesis. So the difference
between the mean is not by chance and it is a marker of a malignant tumor

# Linear Model

TBD

# Bibliography

[1] TBD