---
title: "Foundations of Probability and Statistics, Project"
author: "Emiliano Capasso, Antonello Scarcella, Simone Bellavia"
date: '2023-12-05'
geometry: "left=2cm,right=2cm,top=0.5cm,bottom=1.5cm"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE,
                      tidy.opts=list(width.cutoff = 80),
                      tidy = TRUE)
library(tidyverse) # for tidyverse
library(caret) # for createDataPartition
library(performance) # for compare_performance
library(ggplot2) # for plots
library(reshape2) # for melt function
library(gridExtra) # to arrange plots in a grid 
library(factoextra) # for PCA
library(ggpubr)
library(ggridges)
library(corrplot) # for correlation between variables and PCA
library(psych) # for descriptive statistics
```

# Introduction to Analysis

Breast cancer is one of the most prevalent forms of cancer in women worldwide. According to the World Health Organization, more than 1.7 million 
new cases of breast cancer are diagnosed each year, making it the most common form of cancer among women. Early detection and proper classification 
of the cancer are critical to ensure a positive prognosis and appropriate treatment. 

The **Breast Cancer Wisconsin (Diagnostic) Data Set** provides information on the characteristics of cancer cells found in breast tissue and the 
final diagnosis (malignant or benign). This dataset has been used as a benchmark for many classification algorithms and continues to be a benchmark 
for researchers and developers of artificial intelligence systems in the field of medicine.

This dataset will be used in this project for the analysis of breast cancer. To this end, the project consists of several sections: data exploration, 
descriptive statistical analysis, feature selection with related testing part, and application of the linear model.

This dataset will be used in this project for the analysis of breast cancer. To this end, the report consists of several sections: 

- the first part of the project will be based on **Data Preparation and Cleaning.** We will check the correctness of the type of data available, the presence of missing values and outliers;

- the second part will consist of **Descriptive Statistical Analysis.** Covariances and correlations between features will be checked to give us a better understanding of the nature of the data and its distribution;

- the third part will be based on **Inferential Statistics.** Tests and hypothesis testing will be carried out in order to be able to make considerations about the diagnosis of benign or malignant tumor;

- the fourth part will see the application of the **Linear Model.** The outputs will give more information about the data at hand.

# Data Preparation and Cleaning

## Importing data

The dataset is imported from a CSV file provided by the UCI Repository.

The only feature that identifies the type of diagnosis is represented by, precisely, *diagnosis*. Therefore, being a string, it is converted already as a factor from the import.

```{r}
# import data
data <- read.csv("data.csv",
                    header = TRUE,
                    sep=",",
                    stringsAsFactors = TRUE)
```

A check of the structure of the dataset can be made.

```{r}
# get structure of dataset
str(data)
```

Several considerations can be made about the dataset. It consists of 33 features and 569 observations. Thanks to UCI, it is known
that these features are computed from a digitized image of a fine needle aspirate (FNA) of a breast mass. A fine needle aspiration (FNA) is a type of biopsy. 
It uses a very thin needle and syringe to remove a sample of cells, tissue or fluid from an abnormal area or lump in the body. The sample is then examined under a microscope. 
FNA is also called fine needle aspiration biopsy, or fine needle biopsy. [1] 
In this case, the features describe characteristics of the cell nuclei present in the image. A few of the images can be found at [Web Link](http://www.cs.wisc.edu/~street/images/).

Some information about the features:

1) id: id number;
2) diagnosis (response): the diagnosis of breast tissues (M = malignant, B = benign);

From 3 to 32 ten real-valued features are computed for each cell nucleus:
a) radius (mean of distances from center to points on the perimeter);
b) texture (standard deviation of gray-scale values);
c) perimeter;
d) area;
e) smoothness (local variation in radius lengths);
f) compactness (perimeter^2 / area - 1.0);
g) concavity (severity of concave portions of the contour);
h) concave points (number of concave portions of the contour);
i) symmetry;
j) fractal dimension ("coastline approximation" - 1);

For more in-depth insight, the summary of all attributes and the head of the dataset are presented.

```{r}
# get summary of variables
summary(data)
```

```{r}
# getting the head of dataset
head(data)
```

## Missing Values

It is important to check that the available dataset does not contain missing or null values. For this reason, a spot check is performed.

```{r}
# check for missing values
colSums(is.na(data))
```

There aren't missing values in the considered dataset, except for 32th feature 'X' that is full of NA. 
For this reason, we remove the attribute completely, as having no relevant information is not useful for the analysis.

```{r}
data <- data %>% select(-X)
```

For the same reason, although it does not contain null values, the 'id' attribute is also removed.

```{r}
data <- data %>% select(-id)
```

A check is made on the effective removal of these attributes.

```{r}
colnames(data)
```

# Descriptive Statistical Analysis

The descriptive statistical analysis section aims to explore the properties and relationships among the different variables in the dataset. 
This section will include an analysis of the frequency of cancer diagnoses (malignant or benign), as well as an analysis of the relationship 
between diagnoses and cancer cell attributes. The distribution of attributes and the relationships between them will also be screened, providing 
an overview of the fundamental properties of the dataset. This section will form the basis for the subsequent analysis of the relationships between 
the variables and their importance in breast cancer classification.

## Benignant or Malignant diagnosis

A check is made on the frequency of the two types of breast cancer diagnosis, benign or malignant.

```{r}
ggplot(data, aes(x = diagnosis)) + 
  geom_bar(aes(y = (after_stat(count))/sum(after_stat(count)))) + 
  scale_fill_manual(values = c("#0468BF","#D9A23D")) + 
theme_bw() +
  labs(x = "Diagnosis", y = "Frequency", title = "Frequency of Cancer Stages")
```

It is possible to verify how the frequency of benign tumors is much higher than malignant ones.

The values of the response diagnostic variable are converted to a numerical basis to enable a correlation study: Malignant == 1, Benign == 0.

```{r}
data$diagnosis <- as.numeric(data$diagnosis == "M")

str(data)
```



## Contingecy Tables & Chi-sq Test

Due to the fact that the response variable "diagnosis" it's a categorial one, we can't use
correlation value to analyse the dipendency over the explanatory variables.

We need to create contingency tables and test the indipendence of the variable 
using the Chi-squared test

H0: The two variables are independent.
H1: The two variables relate to each other.

we will keep only the variables which are dependent to the response. Furthermore 
as we need to find which variable is more dependent than the other we create a list
containing all the normalised chi-squared value

```{r}
# function for plotting a dataframe containing variables dependencies with chi-squared values

dependency_list <- function(df) { 
    features_mean <- names(df)[2:11]
    features_se <- names(df)[12:21]
    features_worst <- names(df)[22:31]

    chivaluesN <- c(1)
    indipendentV <- c(FALSE)

    for (x in features_mean) {
        con <- table(cut(df[,x],breaks = 7),df$diagnosis)
        indipendent <- chisq.test(con)$p.value > 0.05
        chivalueN <- round(chisq.test(con)$statistic / length(df$diagnosis),digits = 4)
        indipendentV <- append(indipendentV, indipendent)
        chivaluesN <- append(chivaluesN, chivalueN)
    }

    for (x in features_se) {
        con <- table(cut(df[,x],breaks = 3),df$diagnosis)
        indipendent <- chisq.test(con)$p.value > 0.05
        chivalueN <- round(chisq.test(con)$statistic / length(df$diagnosis),digits = 4)
        indipendentV <- append(indipendentV, indipendent)
        chivaluesN <- append(chivaluesN, chivalueN)
    }

    for (x in features_worst) {
        con <- table(cut(df[,x],breaks = 7),df$diagnosis)
        indipendent <- chisq.test(con)$p.value > 0.05
        chivalueN <- round(chisq.test(con)$statistic / length(df$diagnosis),digits = 4)
        indipendentV <- append(indipendentV, indipendent)
        chivaluesN <- append(chivaluesN, chivalueN)
    }

    features <- names(df)[1:31]
    dv <- data.frame(features,chivaluesN,indipendentV)

    return(dv)
}

dependency_v <- dependency_list(data)
dependency_v <- dependency_v[dependency_v$features != "diagnosis",]
```

we discard all the values which are indipendente so all the TRUE, which correspond with a p-value > 0.05

```{r}
dependency_v <- dependency_v[dependency_v$indipendentV == "FALSE",]
```

of the remains we select those with a chi-squared normalised values > 0.25

```{r}
dependency_v <- dependency_v[dependency_v$chivaluesN > 0.25,]
dependency_v
```

So it is possible to discard: 

- all the variables "*_se";
- texture_*;
- smoothness_*;
- symmerty_*;
- fractal_dimension_*.

On the remaining features, a more in-depth analysis can be conducted

## Correlation map

A correlation map with a heatmap is generated between the variables selected

```{r}

feature_data_matrix <- subset(data, select = dependency_v$features) #%>% select(-diagnosis)
# Calculate the correlation matrix among features
corr_matrix <- cor(feature_data_matrix)

# Calculate the correlation p-value matrix: if p-value is smaller than 0.05, the feature correlation is statistically relavant
p_value_matrix <- matrix(NA, nrow = ncol(feature_data_matrix), ncol = ncol(feature_data_matrix) )

for (i in 1:ncol(feature_data_matrix)) {
  for (j in 1:ncol(feature_data_matrix)) {
    if (i != j) {
      cor_test <- cor.test(feature_data_matrix[,i], feature_data_matrix[,j], conf.level = 0.95)
      p_value_matrix[i,j] <- ifelse(is.na(cor_test$p.value), '', ifelse(cor_test$p.value<0.05, 'S', 'N'))
    }
  }
}

# Melt the correlation matrix into a data frame
melted_matrix <- melt(corr_matrix)
p_melted_matrix <- melt(p_value_matrix)

# Add the correlation coefficient as a label to each tile
melted_matrix$label <- paste0(round(melted_matrix$value, 2), paste0('-', p_melted_matrix$value))

ggplot(data = melted_matrix, aes(x = Var1, y = Var2, fill = value)) + 
  geom_tile(color = "white") +
  geom_text(aes(label = label), color = "black", size = 3) + 
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0,
    limit = c(-1, 1), name = "Correlation") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ggtitle("Correlation Map")
```

It is already possible to find which attributes have a higher correlation with the response variable and 
which, on the other hand, might be discarded by virtue of this criterion.

## Features related to diagnosis

Comparison histograms between features and the distribution of malignant or benign tumor diagnosis are generated. 
These can be conveyed in order to make assertions about their distributions and significance.

```{r}
#features_mean <- names(data)[2:11]

features_mean <- dependency_v$features

plots <- lapply(1:length(features_mean), function(x) {
  g <- ggplot(data, aes_string(x = features_mean[x], 
                              fill = as.factor(data$diagnosis))) +
    geom_histogram(binwidth = (max(data[,features_mean[x]]) - min(data[,features_mean[x]]))/50, 
                  alpha = 0.5, aes(color = as.factor(data$diagnosis))) +
    scale_fill_manual(values = c("#0468BF", "#D9A23D")) +
    scale_color_manual(values = c("#0468BF", "#D9A23D")) +
    ggtitle(features_mean[x]) +
    theme_bw() + 
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(fill = "Diagnosis", color = "Diagnosis")
  return(g)
})

ggarrange(plotlist = plots,
          ncol = 3 , nrow = 2,
          common.legend = T,
          legend = "bottom")


#grid.arrange(grobs = plots, ncol = 2, widths = c(60, 60), height = 60)
```



# Inferential Statistics

The inferential statistical analysis section focuses on using statistical methods to make inferences about the properties of 
populations based on the data in the dataset. This section aims to identify relationships between variables and determine the 
importance of individual variables in breast cancer classification. Hypothesis testing will be used to confirm or reject relationships 
between variables. This section will provide a deeper understanding of the properties of the dataset and their relationship to breast cancer diagnosis.
Finally, regression techniques will be used to determine the relationship between attributes and diagnoses and to identify the most important attributes for tumor classification. 

## Covariance and Correlation

followed the analysis above we can explore more the variables left.

```{r}
cov(data[c("radius_mean","perimeter_mean","area_mean","radius_worst",
"perimeter_worst","area_worst","diagnosis")])

cor(data[c("radius_mean","perimeter_mean","area_mean","radius_worst",
"perimeter_worst","area_worst","diagnosis")])

pairs(data[c("radius_mean","perimeter_mean","area_mean","radius_worst",
"perimeter_worst","area_worst")])
```

from the plot and the correlation values we can see a very strong correlation between all the features
in this table so we can drop them all except for one. We select the feature which has the higher correlation with the response variable diagnosis
so we select the perimeter_worst with a value of 0.6991

we now test the last remained variables

```{r}
cov(data[c("concavity_mean","compactness_mean","concave.points_mean",
"concavity_worst","compactness_worst","concave.points_worst")])

cor(data[c("concavity_mean","compactness_mean","concave.points_mean",
"concavity_worst","compactness_worst","concave.points_worst","diagnosis")])

pairs(data[c("concavity_mean","compactness_mean","concave.points_mean",
"concavity_worst","compactness_worst","concave.points_worst")])
```

from the analysis and the plot we can see that concave.point_worst and concave.points_mean
are strongly correlated so we keep only concave.point_worst which has the higher
correlation with diagnosis (0.6833).
Same goes for concavity and compacteness mean with they respective worst have a correlation
value less than 0.6 but still strongly correlated.
We keep concavity_mean and compactness_mean whose have the higher correlation
with the diagnosis (0.5640,0.3666)

as a summary we plot the correlation matrix of selected features

```{r}
cor(data[c("perimeter_worst","concavity_mean","compactness_mean","concave.points_worst")])

pairs(data[c("perimeter_worst","concavity_mean","compactness_mean","concave.points_worst")])
```

```{r}

data_fs <- data[c("perimeter_worst","concavity_worst","compactness_mean","concave.points_worst","perimeter_worst","diagnosis")]

```

## Test

We want to determine whether the features selected are 
significantly different between healthy (benign) and diseased patients (malignant).

A t-test assigns a “t” test statistic value to each feature. A good feature, 
represented by little to no overlap of the distributions and a large difference in means, would have a high “t” value.

first we divide the dataset
```{r}
mdf <- data[data$diagnosis == 1, ] # group of Malignant tumor
bdf <- data[data$diagnosis == 0, ] # group of Benign tumor
```

```
```{r}
cm <- ggplot(data_fs, aes(x=compactness_mean, group=diagnosis,fill=factor(diagnosis))) +
geom_density(alpha=0.5) +
scale_fill_manual(values = c("#0468BF","#D9A23D")) + 
theme_bw()

pw <- ggplot(data_fs, aes(x=perimeter_worst, group=diagnosis,fill=factor(diagnosis))) +
geom_density(alpha=0.5) +
scale_fill_manual(values = c("#0468BF","#D9A23D")) + 
theme_bw()

cw <- ggplot(data_fs, aes(x=concavity_worst, group=diagnosis,fill=factor(diagnosis))) +
geom_density(alpha=0.5) +
scale_fill_manual(values = c("#0468BF","#D9A23D")) + 
theme_bw()

cp <- ggplot(data_fs, aes(x=concave.points_worst, group=diagnosis,fill=factor(diagnosis))) +
geom_density(alpha=0.5) +
scale_fill_manual(values = c("#0468BF","#D9A23D")) + 
theme_bw()

ggarrange(cm, pw, cw, cp,
          labels = c("A", "B", "C", "D"),
          ncol = 2 , nrow = 2,
          common.legend = T,
          legend = "bottom")
```

```{r}



t.test(mdf$perimeter_worst,bdf$perimeter_worst, alternative="two.sided", var.equal=FALSE,conf.level=0.95)
t.test(mdf$concavity_worst,bdf$concavity_worst, alternative="two.sided", var.equal=FALSE,conf.level=0.95)
t.test(mdf$compactness_mean,bdf$compactness_mean, alternative="two.sided", var.equal=FALSE,conf.level=0.95)
t.test(mdf$concave.points_worst,bdf$concave.points_worst, alternative="two.sided", var.equal=FALSE,conf.level=0.95)

```

from the t value we can say that the better feature which helps us to distinguish
malignant and benign is the concave.point_worst with a t value of 29

# Linear Model

TBD

# Bibliography

[1] https://cancer.ca/en/treatments/tests-and-procedures/fine-needle-aspiration-fna